---
title: "Functions"
author: "Trisha Bhujle"
date: "2025-10-28"
output: html_document
---

```{r} 
# Within a function, you can call on a variable that you have just created within the function. 
calc_shrub_vol <- function(length, width, height) {
  area <- length * width
  volume <- area * height
  return(volume)
}

calc_shrub_vol(0.8, 1.6, 2.0)
# OR calc_shrub_vol(length = 0.8, width = 1.6, height = 2.0)
# OR calc_shrub_vol(height = 2.0, width = 1.6, length = 0.8)
```

You can create default values within the function. 
```{r} 
# Create a default that height is 1.
calc_shrub_vol <- function(length, width, height = 1) {
  area <- length * width
  volume <- area * height
  return(volume)
}

calc_shrub_vol(0.8, 1.6)
# OR calc_shrub_vol(length = 0.8, width = 1.6)
# If I put in a third value it will override the default. For example: calc_shrub_vol(0.8, 1.6, 3.0)
```

Estimate shrub mass. 
```{r} 
calc_shrub_vol <- function(length, width, height) {
  area <- length * width
  volume <- area * height
  return(volume)
}

est_shrub_mass <- function(volume) {
  mass <- 2.65 * volume^0.9
  return(mass)
}

shrub_volume <- calc_shrub_vol(0.8, 1.6, 2.0)
shrub_volume

shrub_mass <- est_shrub_mass(shrub_volume)
shrub_mass
```


Do the above, but in a cleaner way!
```{r} 
library(tidyverse)

calc_shrub_vol <- function(length, width, height) {
  area <- length * width
  volume <- area * height
}

est_shrub_mass <- function(volume) {
  mass <- 2.65 * volume^0.9
}

calc_shrub_vol(0.8, 1.6, 2.0) |>
  est_shrub_mass()
```

Make one function that does the whole thing. 
```{r} 
calc_shrub_vol <- function(length, width, height) {
  area <- length * width
  volume <- area * height
}

est_shrub_mass <- function(volume) {
  mass <- 2.65 * volume^0.9
}

est_shrub_mass_from_raw <- function(length, width, height) {
  volume <- calc_shrub_vol(length, width, height)
  mass <- est_shrub_mass(volume)
  return(mass)
}
```


Practice: Use the function to calculate how many grams there are in 3.75 pounds.
```{r} 
convert_pounds_to_grams <- function(pounds) {
    grams <- 453.6 * pounds
    return(grams)
}

convert_pounds_to_grams(3.75)
```

Practice with the Gapminder dataset. Here is the original: 
```{r}
library(tidyverse)
library(gapminder)

gapminder <- gapminder |> 
  rename("life_exp" = lifeExp, "gdp_per_cap" = gdpPercap)

est <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/countries_estimated.csv')
gapminder_est <- gapminder |> 
  left_join(est)

cntry <- "Belgium"
country_list <- c("Albania", "Canada", "Spain")

dir.create("figures")
dir.create("figures/europe")

country_list <- unique(gapminder$country)

gap_europe <- gapminder_est |> 
  filter(continent == "Europe") |> 
  mutate(gdp_tot = gdp_per_cap * pop)

country_list <- unique(gap_europe$country)

length(country_list)

cntry <- "Albania"

for (cntry in country_list) {

print(str_c("Plotting ", cntry))
  
gap_to_plot <- gap_europe |> 
  filter(country == cntry)

my_plot <- ggplot(data = gap_to_plot, mapping = aes(x = year, y = gdp_tot)) +
  geom_point() +
  labs(title = str_c(cntry, "GDP", sep = " "), 
       subtitle = ifelse(any(gap_to_plot$estimated == "yes"), "Estimated data", "Reported data"))

ggsave(filename = str_c("figures/europe/", cntry, "_gdp_tot.png", sep = ""), plot = my_plot)

}
```


Practice with the Gapminder dataset. Here is us turning it into a function that we can run country by country: 
```{r}
library(tidyverse)
library(gapminder)

gapminder <- gapminder |> 
  rename("life_exp" = lifeExp, "gdp_per_cap" = gdpPercap)

est <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/countries_estimated.csv')
gapminder_est <- gapminder |> 
  left_join(est)

dir.create("figures")
dir.create("figures/europe")

gap_europe <- gapminder_est |> 
  filter(continent == "Europe") |> 
  mutate(gdp_tot = gdp_per_cap * pop)

print_plot <- function(cntry) {

print(str_c("Plotting ", cntry))
  
gap_to_plot <- gap_europe |> 
  filter(country == cntry)

my_plot <- ggplot(data = gap_to_plot, mapping = aes(x = year, y = gdp_tot)) +
  geom_point() +
  labs(title = str_c(cntry, "GDP", sep = " "), 
       subtitle = ifelse(any(gap_to_plot$estimated == "yes"), "Estimated data", "Reported data"))

ggsave(filename = str_c("figures/europe/", cntry, "_gdp_tot.png", sep = ""), plot = my_plot)

}

print_plot("Iceland")
```

Write a for loop to run a function for each of the countries in your country list.
```{r}
library(tidyverse)
library(gapminder)

gapminder <- gapminder |> 
  rename("life_exp" = lifeExp, "gdp_per_cap" = gdpPercap)

est <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/countries_estimated.csv')
gapminder_est <- gapminder |> 
  left_join(est)

dir.create("figures")
dir.create("figures/europe")

gap_europe <- gapminder_est |> 
  filter(continent == "Europe") |> 
  mutate(gdp_tot = gdp_per_cap * pop)

country_list <- unique(gap_europe$country)

print_plot <- function(cntry) {

print(str_c("Plotting ", cntry))
  
gap_to_plot <- gap_europe |> 
  filter(country == cntry)

my_plot <- ggplot(data = gap_to_plot, mapping = aes(x = year, y = gdp_tot)) +
  geom_point() +
  labs(title = str_c(cntry, "GDP", sep = " "), 
       subtitle = ifelse(any(gap_to_plot$estimated == "yes"), "Estimated data", "Reported data"))

ggsave(filename = str_c("figures/europe/", cntry, "_gdp_tot.png", sep = ""), plot = my_plot)

}

for(cntry in country_list) {
  print_plot(cntry)
}
```

Make very minor changes to plot gdp_per_cap instead.
```{r}
library(tidyverse)
library(gapminder)

gapminder <- gapminder |> 
  rename("life_exp" = lifeExp, "gdp_per_cap" = gdpPercap)

est <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/countries_estimated.csv')
gapminder_est <- gapminder |> 
  left_join(est)

dir.create("figures")
dir.create("figures/europe")

gap_europe <- gapminder_est |> 
  filter(continent == "Europe") |> 
  mutate(gdp_tot = gdp_per_cap * pop)

country_list <- unique(gap_europe$country)

print_plot <- function(cntry, stat = "gdp_per_cap") {

print(str_c("Plotting ", cntry))
  
gap_to_plot <- gap_europe |> 
  filter(country == cntry)

my_plot <- ggplot(data = gap_to_plot, mapping = aes(x = year, y = get(stat))) +
  geom_point() +
  labs(title = str_c(cntry, "GDP", sep = " "), 
       subtitle = ifelse(any(gap_to_plot$estimated == "yes"), "Estimated data", "Reported data"),
       y = stat)

ggsave(filename = str_c("figures/europe/", cntry, "_", stat, ".png", sep = ""), plot = my_plot)

}

print_plot("Germany")
```

Change the function to specifiy the file type.
```{r}
library(tidyverse)
library(gapminder)

gapminder <- gapminder |> 
  rename("life_exp" = lifeExp, "gdp_per_cap" = gdpPercap)

est <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/countries_estimated.csv')
gapminder_est <- gapminder |> 
  left_join(est)

dir.create("figures")
dir.create("figures/europe")

gap_europe <- gapminder_est |> 
  filter(continent == "Europe") |> 
  mutate(gdp_tot = gdp_per_cap * pop)

country_list <- unique(gap_europe$country)

print_plot <- function(cntry, stat = "gdp_per_cap", filetype = "pdf") {

gap_to_plot <- gap_europe |> 
  filter(country == cntry)

print(str_c("Plotting ", cntry))
  
my_plot <- ggplot(data = gap_to_plot, mapping = aes(x = year, y = get(stat))) +
  geom_point() +
  labs(title = str_c(cntry, "GDP", sep = " "), 
       subtitle = ifelse(any(gap_to_plot$estimated == "yes"), "Estimated data", "Reported data"),
       y = stat)

ggsave(filename = str_c("figures/europe/", cntry, "_", stat, ".", filetype, sep = ""), plot = my_plot)

}

print_plot("Germany")
# print_plot("Germany", "lifeExp", "jpg")
```





